MSc Thesis

### 2023-01-27 manuscript/main.tex
added Daria's last name
updated Preface, abstract, lay summary

As global agricultural production strains strains under degrading soil fertility and increasing losses due to climate change, research interest in new avenues for production improvement is intensifying. New crop technologies must meet increasing public and regulatory demand for environmental sustainability, encouraging scientists to revisit overlooked or relatively unknown techniques that may unlock productivity gains. One of the promising developments to arise over the past 20 years is the potential of silicon to improve crop plant performance. With benefits to multiple dimensions of crop performance, silicon may be a key tool to guard crop production against uncertain future growing conditions. Our ability to mobilize silicon-based cropping strategies is dependent on a thorough understanding of the ultimate and proximate causes of silicon accumulation, including both the ecological and genetic interactions that can trigger increased uptake. In this thesis, I attempt to extend recent advances in our understanding of silicon ecology in cereal crops, taking an integrative approach towards untangling silicon accumulation in cereal crops. I used a genome-wide association study to identify genetic markers associated with high silicon content, and performed a greenhouse experiment to test for patterns of rapid silicon accumulation in four cereal crops.

@@

Silicon, a naturally occurring element, provides tremendous benefits to plant health, but is not widely utilized in agriculture. One of the main factors limiting it's application in agriculture is our poor understanding of the exact dynamics of how plants absorb and use silicon from the soil. I identified genetic traits that are associated with high silicon content in a relative of bread wheat, as well as demonstrated that cereal crops (e.g. wheat, barley, oats) have the ability to rapidly uptake silicon from the soil. This rapid uptake means that silicon may be a highly effective defence against insect pests. Combining this results with the genetic data, future research can aim towards creating breeding programs to develop cereal crops that can withstand insect damage based on their silicon content. This development could provide an environmentally friendly strategy to maintain output to feed a growing human population.

@@

The research presented in this thesis is original and unpublished. Isaac Peetoom Heida and Dr. Juli Carrillo conceptualized and developed the experiment presented in Chapter 1. Isaac Peetoom Heida, Dr. Juli Carrillo and Dr. Gurcharn Singh Brar conceptualized and developed the experiment presented in Chapter 2.
Isaac Peetoom Heida developed the question and methodology for Chapter 1. Dr. Aaron Beattie, Dr. Mazen Aljarrah, and Dr. Gurcharn Brar provided seeds for the experiment. Isaac Peetoom Heida designed and set up the experiment, processed and analysed the samples, and performed the statistical analysis. Dr. Shaun Barker and the Mineral Deposit Research Unit of the University of British Columbia provided facilities and expertise for the XRF analysis of the tissue samples. (but I paid them, do I still thank?)
For chapter 2, Isaac Peetoom Heida and Dr. Gurcharn Brar designed the experiment. Isaac Peetoom Heida let plot set up and maintenance, with assistance from Grace Wang, Vincent Fetterley, Sara Salad, Katherine Buchanan, Martina Clausen, and Paul Fisher, and Matt Tsuruda. Isaac Peetoom Heida led the sample harvest, processing and analysis. Kelly Wang, Grace Wang, and Chelsea Gowton assisted with the sample harvesting. Dr. Daria [last name], Paul Fisher, Lucas Friesen, Katie Pryer, Dr. Kinga Treder, Carly MacGregor, and Grace Wang all provided invaluable assistance with sample preparation. 
Chapters two and three of this thesis will be submitted to peer-reviewed journals for publication. For the purposes of this manuscript, actions are depicted in the singular first person. 
\section{Introduction}

>>>>

As global agricultural production strains under degrading soil fertility and increasing losses due to climate change, there is growing research interest in new avenues for production improvement. New crop technologies must meet increasing public and regulatory demand for environmental sustainability, encouraging scientists to revisit overlooked or relatively unknown techniques that may unlock productivity gains. One of the promising developments to arise over the past 20 years is the potential of silicon to improve crop plant performance. With benefits to multiple dimensions of crop performance, silicon may be a key tool to guard crop production against uncertain future growing conditions. Our ability to mobilize silicon-based cropping strategies is dependent on a thorough understanding of the ultimate and proximate causes of silicon accumulation, including both the ecological and genetic interactions that can trigger increased uptake. In this thesis, I extend recent advances in our understanding of silicon ecology in cereal crops, testing for the presence of rapid silicification in common canadian crops, as well as using a  genome-wide association study to identify genetic markers associated with high silicon content.

@@

Silicon provides tremendous benefits to plant health, but is not widely utilized in agriculture, despite it being highly abundance and naturally occurring in most soils. One of the main factors limiting it's application in agriculture is our poor understanding of the exact dynamics of how plants absorb and use silicon from the soil. I identified genetic traits associated with high silicon content in \textit{Aegilops tauschii}, a relative of bread wheat, as well as demonstrated that cereal crops (e.g. wheat, barley, oats) have the ability to rapidly uptake silicon from the soil. This rapid uptake means that silicon may be a highly effective defence against insect pests. Combining these results with the genetic data, future research can aim towards creating breeding programs to develop cereal crops that can withstand insect damage based on their silicon content. This development could provide an environmentally friendly strategy to maintain output to feed a growing human population.

@@

The research presented in this thesis is original and unpublished. Isaac Peetoom Heida and Dr. Juli Carrillo, with assistance from Dr. Jean-Thomas Cornelis, conceptualized and developed the experiment presented in Chapter Two. Isaac Peetoom Heida, Dr. Juli Carrillo and Dr. Gurcharn Singh Brar, with input from Dr. Jean-Thomas Cornelis, conceptualized and developed the experiment presented in Chapter Three.
Isaac Peetoom Heida developed the question and methodology for Chapter Two. Dr. Aaron Beattie, Dr. Mazen Aljarrah, and Dr. Gurcharn Brar provided seeds for the experiment. Isaac Peetoom Heida designed and set up the experiment, processed and analysed the samples, and performed the statistical analysis. Dr. Shaun Barker and the Mineral Deposit Research Unit of the University of British Columbia provided facilities and expertise for the XRF analysis of the tissue samples.
For Chapter Three, Isaac Peetoom Heida led plot set up and maintenance, with assistance from Grace Wang, Vincent Fetterley, Sara Salad, Katherine Buchanan, Martina Clausen, and Paul Fisher, and Matt Tsuruda. Isaac Peetoom Heida led the sample harvest, processing and analysis. Kelly Wang, Grace Wang, and Chelsea Gowton assisted with the sample harvesting. Dr. Daria Reshetniak, Paul Fisher, Lucas Friesen, Katie Pryer, Dr. Kinga Treder, Carly MacGregor, and Grace Wang all provided invaluable assistance with sample preparation. 
Chapters Two and Three of this thesis will be submitted to peer-reviewed journals for publication. For the purposes of this manuscript, actions are depicted in the singular first person.

\section{Introduction}
